---
title: "Advanced Spanish Mobility Data Analysis with mobspain"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Spanish Mobility Data Analysis with mobspain}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Advanced Spanish Mobility Data Analysis with mobspain

The `mobspain` package provides a comprehensive toolkit for analyzing Spanish mobility patterns using MITMA (Ministry of Transport, Mobility and Urban Agenda) data. This vignette demonstrates the advanced features introduced in version 0.2.0.

## Setup

```{r setup}
library(mobspain)

# Initialize data directory with version 2 (recommended)
init_data_dir("~/spanish_mobility_data", version = 2)
```

## Enhanced Data Access

### Intelligent Data Retrieval

The package now provides enhanced data access with intelligent caching and version detection:

```{r data-access}
# Get mobility data with auto-detection and caching
mobility_data <- get_enhanced_mobility_data(
  dates = c("2022-01-01", "2022-01-07"),
  data_type = "origin_destination",
  zone_level = "districts",
  use_cache = TRUE
)

# Get data summary
data_summary <- get_data_summary(c("2022-01-01", "2022-01-07"))
print(data_summary$recommended_analysis)
```

### Version-Aware Analysis

```{r version-info}
# Get comprehensive version information
version_info <- get_data_version_info()
print(version_info$comparison)

# Check current version
current_version <- get_current_data_version()
cat("Using data version:", current_version)
```

### Demographic Filtering (Version 2 only)

```{r demographic-filtering}
# Filter by demographics
filtered_data <- get_enhanced_mobility_data(
  dates = c("2022-01-01", "2022-01-07"),
  data_type = "origin_destination",
  zone_level = "districts",
  demographic_filter = list(
    age = c("25-45", "45-65"),
    sex = c("female"),
    income = c("10-15", ">15")
  )
)
```

## Advanced Time Series Analysis

### Seasonal Decomposition and Trend Analysis

```{r time-series}
# Comprehensive time series analysis
ts_analysis <- analyze_mobility_time_series(
  mobility_data,
  temporal_resolution = "daily",
  seasonal_method = "stl",
  detect_anomalies = TRUE,
  trend_analysis = TRUE
)

# Print results
print(ts_analysis)

# Plot time series with anomalies and trends
plot(ts_analysis)
```

### Anomaly Detection

```{r anomaly-detection}
# Access anomaly detection results
anomalies <- ts_analysis$anomaly_detection
cat("Number of anomalies detected:", length(anomalies$indices))

# Get anomaly dates
anomaly_dates <- anomalies$anomaly_dates
print(head(anomaly_dates))
```

## Geospatial Analysis

### Spatial Autocorrelation

```{r spatial-autocorrelation}
# Get spatial zones
districts <- get_enhanced_spatial_zones("districts", version = 2)

# Analyze spatial autocorrelation
spatial_analysis <- analyze_spatial_patterns(
  mobility_data,
  districts,
  analysis_type = "autocorrelation",
  distance_threshold = 50000  # 50km
)

# Print results
print(spatial_analysis)
```

### Hotspot Detection

```{r hotspot-detection}
# Detect spatial hotspots
hotspot_analysis <- analyze_spatial_patterns(
  mobility_data,
  districts,
  analysis_type = "hotspots"
)

# View hotspot summary
print(hotspot_analysis$hotspot_summary)
```

### Accessibility Analysis

```{r accessibility}
# Analyze spatial accessibility
accessibility_analysis <- analyze_spatial_patterns(
  mobility_data,
  districts,
  analysis_type = "accessibility"
)

# Access accessibility metrics
gravity_access <- accessibility_analysis$gravity_accessibility
cumulative_access <- accessibility_analysis$cumulative_accessibility
```

## Interactive Dashboard Creation

### Overview Dashboard

```{r dashboard-overview}
# Create comprehensive overview dashboard
dashboard <- create_mobility_dashboard(
  mobility_data,
  districts,
  dashboard_type = "overview",
  title = "Spanish Mobility Overview - January 2022",
  theme = "default",
  include_filters = TRUE
)

# Save dashboard
create_mobility_dashboard(
  mobility_data,
  districts,
  dashboard_type = "overview",
  title = "Spanish Mobility Overview",
  save_path = "mobility_dashboard.html"
)
```

### Temporal Dashboard

```{r dashboard-temporal}
# Create temporal analysis dashboard
temporal_dashboard <- create_mobility_dashboard(
  mobility_data,
  dashboard_type = "temporal",
  title = "Temporal Mobility Patterns",
  theme = "light"
)
```

### Spatial Dashboard

```{r dashboard-spatial}
# Create spatial analysis dashboard
spatial_dashboard <- create_mobility_dashboard(
  mobility_data,
  districts,
  dashboard_type = "spatial",
  title = "Spatial Mobility Analysis",
  theme = "default"
)
```

## Advanced Data Validation

### Comprehensive Data Validation

```{r data-validation}
# Validate Spanish mobility data
validation_results <- validate_spanish_mobility_data(
  mobility_data,
  version = 2,
  check_completeness = TRUE,
  check_consistency = TRUE,
  check_anomalies = TRUE
)

# View validation summary
print(validation_results$summary)
print(validation_results$recommendations)
```

### Quality Assessment

```{r quality-assessment}
# Check data quality
quality_report <- assess_data_quality(
  mobility_data,
  version = 2,
  spatial_zones = districts
)

# View quality metrics
print(quality_report$metrics)
```

## Machine Learning Integration

### Predictive Analytics

```{r machine-learning}
# Prepare data for machine learning
ml_data <- prepare_ml_data(
  mobility_data,
  target_variable = "n_trips",
  feature_engineering = TRUE
)

# Train prediction model
model <- train_mobility_prediction_model(
  ml_data,
  model_type = "random_forest",
  validation_split = 0.2
)

# Make predictions
predictions <- predict_mobility_patterns(
  model,
  new_data = ml_data$test_data,
  confidence_interval = TRUE
)
```

### Anomaly Detection with ML

```{r ml-anomaly-detection}
# Detect anomalies using machine learning
ml_anomalies <- detect_ml_anomalies(
  mobility_data,
  method = "isolation_forest",
  contamination = 0.1
)

# View anomaly summary
print(ml_anomalies$summary)
```

## Performance Optimization

### Parallel Processing

```{r parallel-processing}
# Enable parallel processing for large datasets
library(parallel)

# Process multiple date ranges in parallel
date_ranges <- list(
  c("2022-01-01", "2022-01-07"),
  c("2022-01-08", "2022-01-14"),
  c("2022-01-15", "2022-01-21")
)

# Parallel data retrieval
results <- mclapply(date_ranges, function(dates) {
  get_enhanced_mobility_data(
    dates = dates,
    data_type = "origin_destination",
    zone_level = "districts"
  )
}, mc.cores = 2)
```

### Caching Strategies

```{r caching}
# Check cache status
cache_status <- check_cache_status()
print(cache_status)

# Optimize cache
optimize_cache(
  max_size_gb = 10,
  cleanup_old_files = TRUE
)
```

## Advanced Visualization

### Interactive Flow Maps

```{r interactive-flows}
# Create interactive flow map
flow_map <- create_interactive_flow_map(
  mobility_data,
  districts,
  flow_threshold = 1000,
  animation = TRUE,
  time_column = "hour"
)

# Save interactive map
save_interactive_map(flow_map, "flow_map.html")
```

### Enhanced Choropleth Maps

```{r choropleth-maps}
# Create enhanced choropleth map
choropleth <- create_enhanced_choropleth_map(
  districts,
  value_column = "total_trips",
  color_palette = "viridis",
  interactive = TRUE,
  popup_info = c("name", "total_trips", "population")
)
```

## Integration with External Data

### Joining with Census Data

```{r census-integration}
# Join with external census data
enhanced_data <- join_with_census_data(
  districts,
  census_variables = c("population", "income", "age_distribution"),
  census_year = 2021
)
```

### Weather Data Integration

```{r weather-integration}
# Join with weather data
weather_enhanced <- join_with_weather_data(
  mobility_data,
  weather_variables = c("temperature", "precipitation", "wind_speed"),
  weather_source = "meteogalicia"
)
```

## Best Practices and Tips

### Data Size Management

```{r data-management}
# Estimate data size before download
size_estimate <- estimate_download_size(
  dates = c("2022-01-01", "2022-12-31"),
  data_type = "origin_destination",
  zone_level = "districts"
)

cat("Estimated download size:", size_estimate$total_gb, "GB")

# Use data subsetting for large analyses
subset_data <- get_enhanced_mobility_data(
  dates = c("2022-01-01", "2022-01-07"),
  data_type = "origin_destination",
  zone_level = "districts",
  aggregate_temporal = TRUE,  # Aggregate by day
  demographic_filter = list(age = c("25-45"))  # Filter demographics
)
```

### Error Handling

```{r error-handling}
# Robust data retrieval with error handling
tryCatch({
  mobility_data <- get_enhanced_mobility_data(
    dates = c("2022-01-01", "2022-01-07"),
    data_type = "origin_destination",
    zone_level = "districts"
  )
}, error = function(e) {
  cat("Error occurred:", e$message)
  # Fallback to cached data
  mobility_data <- get_cached_data(
    dates = c("2022-01-01", "2022-01-07"),
    data_type = "origin_destination"
  )
})
```

## Conclusion

The `mobspain` package version 0.2.0 provides a comprehensive toolkit for advanced Spanish mobility data analysis. Key improvements include:

1. **Enhanced Data Access**: Intelligent caching, version detection, and demographic filtering
2. **Advanced Analytics**: Time series analysis, geospatial analysis, and machine learning integration
3. **Interactive Dashboards**: Comprehensive dashboard creation with multiple themes and layouts
4. **Data Quality**: Comprehensive validation and quality assessment tools
5. **Performance**: Parallel processing and optimized caching strategies

For more detailed examples and use cases, see the additional vignettes:
- Time Series Analysis (`vignette("time-series", package = "mobspain")`)
- Geospatial Analysis (`vignette("geospatial", package = "mobspain")`)
- Dashboard Creation (`vignette("dashboards", package = "mobspain")`)
- Machine Learning (`vignette("machine-learning", package = "mobspain")`)
