---
title: "Getting Started with mobspain"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mobspain}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Prevent actual data downloads during package build
)
```

```{r library}
library(mobspain)
library(dplyr)
```

## Introduction

The `mobspain` package provides a comprehensive toolkit for analyzing Spanish mobility patterns using MITMA (Ministry of Transport, Mobility and Urban Agenda) data. This vignette demonstrates the main functionality.

### About MITMA Data

The MITMA mobility data is derived from mobile phone analytics and provides origin-destination (OD) matrices representing population movements across Spain. Key characteristics:

- **Spatial Coverage**: Three administrative levels available
  - Districts (`dist`): ~3,909 zones (finest granularity)
  - Municipalities (`muni`): ~8,131 zones  
  - Large Urban Areas (`lua`): ~85 zones (metropolitan areas)
- **Temporal Coverage**: Data available from February 14, 2020 onwards
- **Time Resolution**: Hourly data (0-23 hours) and daily aggregations
- **Data Source**: Aggregated and anonymized mobile phone positioning data
- **Methodology**: Based on official Spanish transport ministry standards
- **Reference**: [MITMA Methodological Report v8](https://www.transportes.gob.es/recursos_mfom/paginabasica/recursos/a3_informe_metodologico_estudio_movilidad_mitms_v8.pdf)

## Setup

First, initialize a data directory for storing MITMA data:

```{r setup_data}
# Set up data directory with version options
init_data_dir(
  path = "~/spanish_mobility_data",  # Custom directory path
  version = 2                        # Version 1 (2020-2021) or 2 (2022+, recommended)
)

# Configure package with specific options
configure_mobspain(
  parallel = TRUE,              # Enable parallel processing
  n_cores = 4,                  # Number of CPU cores (default: detectCores()-1)
  cache_enabled = TRUE,         # Enable data caching
  cache_max_size = 500,         # Cache size in MB
  cache_max_age_days = 7        # Cache expiration in days
)

# Check package status and configuration
status <- mobspain_status()
print("Package configuration:")
print(status)

# Get optimal parameters for different analysis types
optimal_params <- get_optimal_parameters(
  analysis_type = "exploratory",  # Options: "exploratory", "production", "research"
  data_size = "medium"            # Options: "small", "medium", "large"
)
print("Recommended parameters for exploratory analysis:")
print(optimal_params)
```

## Working with Sample Data

The package includes sample data for testing and learning:

```{r sample_data}
# Load sample zones
data(sample_zones)
print(sample_zones)

# Create spatial index
zone_index <- create_zone_index(sample_zones)
```

## Real Data Analysis

### Getting Spatial Zones

Download Spanish administrative boundaries that correspond to MITMA data zones:

```{r spatial_zones}
# Get districts (finest spatial resolution) with specific parameters
districts <- get_spatial_zones(
  level = "dist",              # Level options: "dist", "muni", "lua"
  version = 2,                 # Data version: 1 (2020-2021) or 2 (2022+)
  update = FALSE,              # Force update from server
  cache = TRUE                 # Use cached data if available
)  # Results: ~3,909 zones

# Get municipalities (standard administrative units)
municipalities <- get_spatial_zones(
  level = "muni",              # Municipality level
  version = 2
)  # Results: ~8,131 zones

# Get large urban areas (metropolitan areas)
large_urban_areas <- get_spatial_zones(
  level = "lua",               # Large Urban Areas level
  version = 2
)  # Results: ~85 zones

# Check the structure and contents of spatial data
print(paste("Districts loaded:", nrow(districts), "zones"))
print(paste("Columns available:", paste(names(districts), collapse = ", ")))
print(paste("Spatial zones with area data:", sum(!is.na(districts$area_km2))))

# Access key columns: districts$id, districts$name, districts$geometry, districts$area_km2
head(districts[, c("id", "name", "area_km2")])
```

### Mobility Data Retrieval

Get real mobility matrices from MITMA. The data represents aggregated trips between spatial zones:

```{r mobility_data}
# Get mobility data with comprehensive parameters
mobility_data <- get_mobility_matrix(
  dates = c("2023-01-01", "2023-01-07"),  # Date range (YYYY-MM-DD format)
  level = "dist",                          # Spatial level: "dist", "muni", "lua"
  version = 2,                             # Data version: 1 or 2
  time_window = NULL,                      # Time range: c(start_hour, end_hour), e.g., c(7, 9)
  filter_weekends = FALSE,                 # Exclude weekends: TRUE/FALSE
  validate = TRUE                          # Validate data quality: TRUE/FALSE
)
# Returns data.frame with columns: id_origin, id_destination, date, n_trips, distance_km

# Get morning rush hour data (7-9 AM) for commuting analysis
commute_data <- get_mobility_matrix(
  dates = c("2023-03-01", "2023-03-07"), 
  level = "dist",
  time_window = c(7, 9),          # Morning commute hours (7 AM to 9 AM)
  version = 2
)

# Evening rush hour data (17-19 PM)
evening_data <- get_mobility_matrix(
  dates = c("2023-03-01", "2023-03-07"),
  level = "dist", 
  time_window = c(17, 19),        # Evening commute hours (5 PM to 7 PM)
  version = 2
)

# Weekend-only data for leisure analysis
weekend_data <- get_mobility_matrix(
  dates = c("2023-01-07", "2023-01-08"),  # Saturday-Sunday
  level = "muni",                          # Municipality level for faster processing
  version = 2
)

# Check data structure and key columns
print("Mobility data structure:")
print(str(mobility_data))
print("Key columns:", paste(names(mobility_data), collapse = ", "))
print("Date range:", range(mobility_data$date))
print("Total rows:", nrow(mobility_data))
```

## Analytics

### Understanding MITMA Mobility Data

The mobility matrices represent estimated trips between zones based on mobile phone positioning:

```{r data_exploration}
# Examine the structure of mobility data with specific columns
print("Mobility data structure:")
print(head(mobility_data))
print("Column types:")
print(sapply(mobility_data, class))

# Validate data quality with detailed parameters
quality_report <- validate_mitma_data(
  mobility_data,
  min_trips = 1,                    # Minimum trip threshold
  check_symmetry = TRUE,            # Check origin-destination symmetry
  check_dates = TRUE,               # Validate date formats
  verbose = TRUE                    # Print detailed validation info
)
print("Data quality summary:")
print(quality_report)
# Access results: quality_report$is_valid, quality_report$issues, quality_report$summary

# Check for Spanish holidays with specific parameters
holiday_check <- check_spanish_holidays(
  dates = unique(mobility_data$date),
  include_regional = TRUE,          # Include regional holidays
  buffer_days = 1                   # Check days around holidays
)
holidays_found <- holiday_check[holiday_check$is_likely_holiday, ]
if(nrow(holidays_found) > 0) {
  print("Holidays found in data period:")
  print(holidays_found[, c("date", "holiday_name", "holiday_type")])
}

# Comprehensive data coverage statistics
total_trips <- sum(mobility_data$n_trips, na.rm = TRUE)
unique_origins <- length(unique(mobility_data$id_origin))
unique_destinations <- length(unique(mobility_data$id_destination))
date_range <- range(mobility_data$date)
avg_daily_trips <- total_trips / length(unique(mobility_data$date))

print("Data coverage statistics:")
print(paste("Total trips recorded:", format(total_trips, big.mark = ",")))
print(paste("Unique origin zones:", unique_origins))
print(paste("Unique destination zones:", unique_destinations))
print(paste("Date range:", date_range[1], "to", date_range[2]))
print(paste("Average daily trips:", format(round(avg_daily_trips), big.mark = ",")))
```

### Containment Analysis

Self-containment measures the proportion of trips that stay within the same zone (internal mobility vs. external trips):

```{r containment}
# Calculate containment indices with specific parameters
containment <- calculate_containment(
  mobility_data,
  min_trips = 10,                   # Minimum trips threshold for inclusion
  normalize = TRUE,                 # Normalize by total trips
  include_totals = TRUE             # Include total trip counts
)
# Returns data.frame with: id_origin, containment, total_trips, internal_trips, external_trips

# Display most self-contained zones with specific sorting
print("Top 10 most self-contained zones:")
top_contained <- head(containment[order(-containment$containment), ], 10)
print(top_contained[, c("id_origin", "containment", "total_trips", "internal_trips")])

# Summary statistics with quantiles
containment_stats <- summary(containment$containment)
print("Containment index statistics:")
print(containment_stats)

# Additional analysis
high_containment <- sum(containment$containment > 0.5, na.rm = TRUE)
very_high_containment <- sum(containment$containment > 0.8, na.rm = TRUE)
print(paste("Zones with >50% containment:", high_containment))
print(paste("Zones with >80% containment:", very_high_containment))
print(paste("Mean containment index:", round(mean(containment$containment, na.rm = TRUE), 3)))
print(paste("Median containment index:", round(median(containment$containment, na.rm = TRUE), 3)))
```

### Advanced Mobility Indicators

Compute comprehensive mobility metrics including connectivity, flow balance, and spatial characteristics:

```{r indicators}
# Calculate comprehensive mobility indicators with parameters
indicators <- calculate_mobility_indicators(
  mobility_data, 
  districts,
  include_distance = TRUE,          # Include distance-based metrics
  normalize_by_population = FALSE,  # Normalize by population (if available)
  min_trips = 5,                    # Minimum trips threshold
  include_centrality = TRUE         # Include network centrality measures
)
# Returns data.frame with columns: id_origin, total_outflow, total_inflow, containment, 
# connectivity_index, net_flow, avg_distance, centrality_score

# Display key indicators with specific columns
print("Sample of mobility indicators:")
indicator_cols <- c("id_origin", "total_outflow", "total_inflow", "containment", 
                   "connectivity_index", "net_flow", "avg_distance")
print(head(indicators[, indicator_cols]))

# Analyze mobility patterns with detailed statistics
outflow_zones <- sum(indicators$net_flow < 0, na.rm = TRUE)
inflow_zones <- sum(indicators$net_flow > 0, na.rm = TRUE)
balanced_zones <- sum(abs(indicators$net_flow) < 100, na.rm = TRUE)

print("Mobility pattern analysis:")
print(paste("Zones with net outflow (job/service sources):", outflow_zones))
print(paste("Zones with net inflow (employment/service centers):", inflow_zones))
print(paste("Balanced zones (inflow ≈ outflow):", balanced_zones))

# Connectivity analysis
connectivity_stats <- summary(indicators$connectivity_index)
print("Connectivity index statistics:")
print(connectivity_stats)

# Distance analysis (if available)
if("avg_distance" %in% names(indicators)) {
  distance_stats <- summary(indicators$avg_distance)
  print("Average distance statistics (km):")
  print(distance_stats)
}
```

### Anomaly Detection

Detect unusual mobility patterns that deviate from normal behavior (useful for identifying special events, holidays, disruptions):

```{r anomalies}
# Detect anomalies with comprehensive parameters
anomalies <- detect_mobility_anomalies(
  mobility_data, 
  method = "zscore",                # Methods: "zscore", "iqr", "isolation"
  threshold = 2.5,                  # Z-score threshold (2.5 = 98.7th percentile)
  by_weekday = TRUE,                # Separate analysis by weekday/weekend
  min_trips = 100,                  # Minimum daily trips for analysis
  rolling_window = 7                # Rolling window for trend analysis
)
# Returns data.frame with: date, total_trips, z_score, is_anomaly, weekday, is_weekend

# Show detected anomalies with detailed information
print("Detected mobility anomalies:")
if(any(anomalies$is_anomaly)) {
  anomaly_dates <- anomalies[anomalies$is_anomaly, ]
  anomaly_cols <- c("date", "total_trips", "z_score", "is_anomaly", "weekday", "is_weekend")
  print(anomaly_dates[, anomaly_cols])
  
  # Statistical summary of anomalies
  print("Anomaly statistics:")
  print(paste("Total anomalies detected:", sum(anomalies$is_anomaly)))
  print(paste("Anomaly rate:", round(mean(anomalies$is_anomaly) * 100, 1), "%"))
  print(paste("Max z-score:", round(max(anomalies$z_score, na.rm = TRUE), 2)))
  print(paste("Min z-score:", round(min(anomalies$z_score, na.rm = TRUE), 2)))
  
  # Check if anomalies correspond to holidays
  anomaly_holidays <- check_spanish_holidays(
    dates = anomaly_dates$date,
    include_regional = TRUE
  )
  holiday_anomalies <- anomaly_holidays[anomaly_holidays$is_likely_holiday, ]
  
  if(nrow(holiday_anomalies) > 0) {
    print("Anomalies corresponding to holidays:")
    print(holiday_anomalies[, c("date", "holiday_name", "holiday_type")])
  }
} else {
  print("No significant anomalies detected in this period")
}

# Compare weekend vs weekday patterns with detailed analysis
print("Weekend vs Weekday mobility patterns:")
weekend_summary <- anomalies %>%
  dplyr::group_by(is_weekend) %>%
  dplyr::summarise(
    count = n(),
    avg_trips = mean(total_trips, na.rm = TRUE),
    median_trips = median(total_trips, na.rm = TRUE),
    anomaly_rate = mean(is_anomaly, na.rm = TRUE) * 100,
    avg_z_score = mean(abs(z_score), na.rm = TRUE),
    .groups = "drop"
  )
print(weekend_summary)
```

### Distance-Decay Modeling

Model how mobility flows decrease with distance (fundamental principle in spatial interaction modeling):

```{r distance_decay}
# Calculate distance-decay relationship with detailed parameters
decay_model <- calculate_distance_decay(
  mobility_data, 
  districts,
  max_distance = 500,               # Maximum distance in km
  distance_bins = 25,               # Distance bin size in km
  min_trips_per_bin = 50,           # Minimum trips per distance bin
  method = "power_law",             # Model type: "power_law", "exponential", "linear"
  weight_by_trips = TRUE            # Weight regression by trip volume
)
# Returns list with: model, r_squared, decay_exponent, coefficients, distance_bins, model_summary

# Display comprehensive model results
print("Distance-decay model summary:")
print(decay_model$model_summary)

# Key model statistics
print("Model performance metrics:")
print(paste("R-squared:", round(decay_model$r_squared, 3)))
print(paste("Distance decay exponent:", round(decay_model$decay_exponent, 3)))
print(paste("Model type:", decay_model$model_type))

# Access specific results
coefficients <- decay_model$coefficients
print("Model coefficients:")
print(coefficients)

# Interpretation guidelines
exponent <- decay_model$decay_exponent
print("Interpretation:")
if(abs(exponent) < 0.5) {
  print("Weak distance decay - mobility remains high over long distances")
} else if(abs(exponent) < 1.5) {
  print("Moderate distance decay - typical urban mobility pattern")
} else {
  print("Strong distance decay - highly localized mobility")
}

# Distance bin analysis
if("distance_bins" %in% names(decay_model)) {
  print("Distance bin statistics:")
  print(head(decay_model$distance_bins))
}
```

## Visualization

### Time Series Plots

Visualize mobility trends over time:

```{r daily_mobility}
# Create daily mobility plot with styling options
daily_plot <- plot_daily_mobility(
  mobility_data,
  group_by = "weekday",            # Options: "weekday", "date", "none"
  smooth = TRUE,                   # Add trend line
  color_by = "day_type",           # Color by weekend/weekday
  title = "Daily Mobility Patterns",
  y_axis_label = "Total Trips",
  date_labels = "%b %d"            # Date format
)
print(daily_plot)

# Alternative: Simple time series plot
simple_plot <- plot_daily_mobility(
  mobility_data,
  group_by = "date",
  smooth = FALSE
)
print(simple_plot)
```

### Interactive Flow Maps

Create interactive maps of mobility flows:

```{r flow_maps}
# Create interactive flow map with comprehensive styling
flow_map <- create_flow_map(
  districts, 
  mobility_data, 
  min_flow = 500,                  # Minimum flow threshold
  max_flows = 100,                 # Maximum number of flows to display
  line_color = "blue",             # Flow line color
  line_width_scale = 1.5,          # Line width scaling factor
  line_opacity = 0.7,              # Line transparency (0-1)
  background_map = "cartodb",      # Options: "osm", "cartodb", "stamen"
  zoom_level = 6,                  # Initial zoom level
  cluster_origins = TRUE           # Cluster origin points
)
# Returns leaflet map object
flow_map

# Alternative: Simple flow map
simple_flow_map <- create_flow_map(
  districts, 
  mobility_data, 
  min_flow = 1000,
  background_map = "osm"
)
simple_flow_map
```

### Choropleth Maps

Create maps showing spatial indicators:

```{r choropleth}
# Create choropleth map with detailed styling options
choropleth <- create_choropleth_map(
  districts, 
  indicators, 
  variable = "containment",         # Column name to visualize
  color_palette = "viridis",        # Options: "viridis", "plasma", "blues", "reds"
  legend_title = "Self-containment Index",
  breaks = 5,                       # Number of color breaks
  na_color = "grey",                # Color for missing values
  border_color = "white",           # Zone border color
  border_weight = 0.5,              # Border line weight
  popup_columns = c("name", "containment", "total_trips")  # Popup information
)
# Returns leaflet map object
choropleth

# Alternative: Different variable and styling
connectivity_map <- create_choropleth_map(
  districts,
  indicators,
  variable = "connectivity_index",
  color_palette = "plasma",
  legend_title = "Connectivity Index",
  breaks = 7
)
connectivity_map
```

### Heatmaps

Create flow matrix heatmaps:

```{r heatmaps}
# Create mobility heatmap with styling parameters
heatmap <- plot_mobility_heatmap(
  mobility_data, 
  top_n = 20,                      # Top N zones to include
  cluster_rows = TRUE,             # Cluster similar zones
  cluster_cols = TRUE,             # Cluster similar time periods
  color_scale = "log10",           # Options: "linear", "log10", "sqrt"
  color_palette = "viridis",       # Color palette
  show_values = FALSE,             # Show numeric values in cells
  title = "Mobility Flow Heatmap",
  x_axis_label = "Destination Zones",
  y_axis_label = "Origin Zones"
)
print(heatmap)

# Alternative: Simple heatmap with different parameters
simple_heatmap <- plot_mobility_heatmap(
  mobility_data,
  top_n = 15,
  cluster_rows = FALSE,
  color_scale = "linear",
  color_palette = "blues"
)
print(simple_heatmap)
```

### Distance-Decay Plots

Visualize distance-decay relationships:

```{r distance_plots}
# Create distance-decay visualization with styling options
decay_plot <- plot_distance_decay(
  decay_model,
  point_color = "blue",            # Data point color
  line_color = "red",              # Regression line color
  point_size = 2,                  # Point size
  line_width = 1.5,                # Line width
  add_r_squared = TRUE,            # Show R-squared on plot
  log_scale = TRUE,                # Use log scale for axes
  title = "Distance-Decay Relationship",
  x_axis_label = "Distance (km)",
  y_axis_label = "Trip Frequency"
)
print(decay_plot)

# Alternative: Linear scale plot
linear_plot <- plot_distance_decay(
  decay_model,
  log_scale = FALSE,
  point_color = "darkgreen",
  line_color = "orange"
)
print(linear_plot)
```

## Configuration

The package supports advanced configuration:

```{r config}
# Configure package with comprehensive options
configure_mobspain(
  parallel = TRUE,                  # Enable parallel processing
  n_cores = 4,                      # Number of CPU cores
  cache_enabled = TRUE,             # Enable data caching
  cache_max_size = 1000,            # Cache size in MB
  cache_max_age_days = 14,          # Cache expiration (days)
  data_source = "csv",              # Options: "csv", "duckdb"
  validate_data = TRUE,             # Validate data on load
  verbose = TRUE                    # Print status messages
)

# Check detailed configuration status
status <- mobspain_status()
print("Package configuration:")
print(status)
# Access specific settings: status$parallel, status$cache_enabled, status$data_dir

# Cache management
cache_info <- get_cache_info()
print("Cache information:")
print(cache_info)

# Clean old cache files
cleanup_cache(max_age_days = 30)
```

## Conclusion

The `mobspain` package provides a complete toolkit for Spanish mobility analysis, from data retrieval to advanced analytics and visualization. 

### Key Methodological Notes

When working with MITMA data, consider:

- **Data Quality**: Mobile phone data represents a sample of the population; results should be interpreted as relative patterns rather than absolute counts
- **Temporal Patterns**: Weekdays vs. weekends show different mobility patterns; holidays and special events may create anomalies
- **Spatial Resolution**: District-level data provides fine detail but requires more computational resources; municipality or LUA levels may be more appropriate for regional studies
- **Privacy Protection**: All data is aggregated and anonymized according to Spanish data protection standards

### Further Resources

- **MITMA Methodology**: [Official methodological report](https://www.transportes.gob.es/recursos_mfom/paginabasica/recursos/a3_informe_metodologico_estudio_movilidad_mitms_v8.pdf)
- **Package Documentation**: `?mobspain` or `help(package = "mobspain")`
- **Example Scripts**: See `complete_working_example.R` and `simple_mobspain_example.R` in the package repository

For more information, see the package documentation and example scripts.
