---
title: "Getting Started with mobspain"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mobspain}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Prevent actual data downloads during package build
)
```

```{r library}
library(mobspain)
```

## Introduction

The `mobspain` package provides a comprehensive toolkit for analyzing Spanish mobility patterns using MITMA (Ministry of Transport, Mobility and Urban Agenda) data. This vignette demonstrates the main functionality.

## Setup

First, initialize a data directory for storing MITMA data:

```{r setup_data}
# Set up data directory
init_data_dir("~/spanish_mobility_data")

# Check package status
mobspain_status()
```

## Working with Sample Data

The package includes sample data for testing and learning:

```{r sample_data}
# Load sample zones
data(sample_zones)
print(sample_zones)

# Create spatial index
zone_index <- create_zone_index(sample_zones)
```

## Real Data Analysis

### Getting Spatial Zones

Download real Spanish administrative zones:

```{r spatial_zones}
# Get districts (around 3,909 zones)
districts <- get_spatial_zones("dist")

# Get municipalities
municipalities <- get_spatial_zones("muni")
```

### Mobility Data Retrieval

Get real mobility matrices from MITMA:

```{r mobility_data}
# Get mobility data for a week
mobility_data <- get_mobility_matrix(
  dates = c("2023-01-01", "2023-01-07"),
  level = "dist"
)

# Get morning commute data
commute_data <- get_mobility_matrix(
  dates = c("2023-03-01", "2023-03-07"), 
  level = "dist",
  time_window = c(7, 9)  # 7-9 AM
)
```

## Analytics

### Containment Analysis

Calculate self-containment indices:

```{r containment}
containment <- calculate_containment(mobility_data)
head(containment[order(-containment$containment), ])
```

### Advanced Mobility Indicators

Compute comprehensive mobility metrics:

```{r indicators}
indicators <- calculate_mobility_indicators(mobility_data, districts)
print(indicators)
```

### Anomaly Detection

Detect unusual mobility patterns:

```{r anomalies}
anomalies <- detect_mobility_anomalies(mobility_data, method = "zscore")
```

### Distance-Decay Modeling

Model the relationship between distance and mobility flows:

```{r distance_decay}
decay_model <- calculate_distance_decay(mobility_data, districts)
```

## Visualization

### Time Series Plots

Visualize mobility trends over time:

```{r daily_mobility}
daily_plot <- plot_daily_mobility(mobility_data)
print(daily_plot)
```

### Interactive Flow Maps

Create interactive maps of mobility flows:

```{r flow_maps}
# Create flow map with minimum flow threshold
flow_map <- create_flow_map(districts, mobility_data, min_flow = 500)
flow_map
```

### Choropleth Maps

Create maps showing spatial indicators:

```{r choropleth}
choropleth <- create_choropleth_map(
  districts, 
  indicators, 
  variable = "containment",
  title = "Self-Containment by District"
)
choropleth
```

### Heatmaps

Create flow matrix heatmaps:

```{r heatmaps}
heatmap <- plot_mobility_heatmap(mobility_data, top_n = 20)
print(heatmap)
```

### Distance-Decay Plots

Visualize distance-decay relationships:

```{r distance_plots}
decay_plot <- plot_distance_decay(decay_model)
print(decay_plot)
```

## Configuration

The package supports advanced configuration:

```{r config}
# Enable parallel processing
configure_mobspain(parallel = TRUE, n_cores = 4)

# Check configuration
mobspain_status()
```

## Conclusion

The `mobspain` package provides a complete toolkit for Spanish mobility analysis, from data retrieval to advanced analytics and visualization. For more information, see the package documentation and example scripts.
