---
title: "Getting Started with mobspain"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mobspain}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Prevent actual data downloads during package build
)
```

```{r library}
library(mobspain)
```

## Introduction

The `mobspain` package provides a comprehensive toolkit for analyzing Spanish mobility patterns using MITMA (Ministry of Transport, Mobility and Urban Agenda) data. This vignette demonstrates the main functionality.

### About MITMA Data

The MITMA mobility data is derived from mobile phone analytics and provides origin-destination (OD) matrices representing population movements across Spain. Key characteristics:

- **Spatial Coverage**: Three administrative levels available
  - Districts (`dist`): ~3,909 zones (finest granularity)
  - Municipalities (`muni`): ~8,131 zones  
  - Large Urban Areas (`lua`): ~85 zones (metropolitan areas)
- **Temporal Coverage**: Data available from February 14, 2020 onwards
- **Time Resolution**: Hourly data (0-23 hours) and daily aggregations
- **Data Source**: Aggregated and anonymized mobile phone positioning data
- **Methodology**: Based on official Spanish transport ministry standards
- **Reference**: [MITMA Methodological Report v8](https://www.transportes.gob.es/recursos_mfom/paginabasica/recursos/a3_informe_metodologico_estudio_movilidad_mitms_v8.pdf)

## Setup

First, initialize a data directory for storing MITMA data:

```{r setup_data}
# Set up data directory
init_data_dir("~/spanish_mobility_data")

# Check package status
mobspain_status()

# Get optimal parameters for exploratory analysis
optimal_params <- get_optimal_parameters("exploratory", "medium")
print("Recommended parameters for exploratory analysis:")
print(optimal_params)
```

## Working with Sample Data

The package includes sample data for testing and learning:

```{r sample_data}
# Load sample zones
data(sample_zones)
print(sample_zones)

# Create spatial index
zone_index <- create_zone_index(sample_zones)
```

## Real Data Analysis

### Getting Spatial Zones

Download Spanish administrative boundaries that correspond to MITMA data zones:

```{r spatial_zones}
# Get districts (finest spatial resolution)
# These are sub-municipal divisions used by MITMA
districts <- get_spatial_zones("dist")  # ~3,909 zones

# Get municipalities (standard administrative units)
# These correspond to official Spanish municipalities
municipalities <- get_spatial_zones("muni")  # ~8,131 zones

# Get large urban areas (metropolitan areas)
# These represent major urban agglomerations
large_urban_areas <- get_spatial_zones("lua")  # ~85 zones

# Check the structure of spatial data
print(paste("Districts loaded:", nrow(districts), "zones"))
print(paste("Area calculated for:", sum(!is.na(districts$area_km2)), "zones"))
```

### Mobility Data Retrieval

Get real mobility matrices from MITMA. The data represents aggregated trips between spatial zones:

```{r mobility_data}
# Get mobility data for a week (districts level)
# Note: Dates should be >= 2020-02-14 when MITMA data starts
mobility_data <- get_mobility_matrix(
  dates = c("2023-01-01", "2023-01-07"),
  level = "dist"  # Districts: ~3,909 zones
)

# Get morning rush hour data (7-9 AM)
# This captures commuting patterns
commute_data <- get_mobility_matrix(
  dates = c("2023-03-01", "2023-03-07"), 
  level = "dist",
  time_window = c(7, 9)  # Morning commute hours
)

# Alternative: Get municipality-level data (fewer zones, faster processing)
muni_mobility <- get_mobility_matrix(
  dates = c("2023-01-01", "2023-01-03"),
  level = "muni"  # Municipalities: ~8,131 zones
)
```

## Analytics

### Understanding MITMA Mobility Data

The mobility matrices represent estimated trips between zones based on mobile phone positioning:

```{r data_exploration}
# Examine the structure of mobility data
print("Mobility data structure:")
print(head(mobility_data))

# Validate data quality
quality_report <- validate_mitma_data(mobility_data)
print("Data quality summary:")
print(quality_report)

# Check for holidays that might affect patterns
holiday_check <- check_spanish_holidays(unique(mobility_data$date))
holidays_found <- holiday_check[holiday_check$is_likely_holiday, ]
if(nrow(holidays_found) > 0) {
  print("Holidays found in data period:")
  print(holidays_found)
}

# Basic data coverage statistics
print(paste("Total trips recorded:", sum(mobility_data$n_trips)))
print(paste("Number of origin zones:", length(unique(mobility_data$id_origin))))
print(paste("Number of destination zones:", length(unique(mobility_data$id_destination))))
print(paste("Date range:", min(mobility_data$date), "to", max(mobility_data$date)))
```

### Containment Analysis

Self-containment measures the proportion of trips that stay within the same zone (internal mobility vs. external trips):

```{r containment}
# Calculate containment indices for all zones
containment <- calculate_containment(mobility_data)

# Display most self-contained zones
print("Top 10 most self-contained zones:")
top_contained <- head(containment[order(-containment$containment), ], 10)
print(top_contained)

# Summary statistics
print(paste("Mean containment index:", round(mean(containment$containment, na.rm = TRUE), 3)))
print(paste("Zones with >50% containment:", sum(containment$containment > 0.5, na.rm = TRUE)))
```

### Advanced Mobility Indicators

Compute comprehensive mobility metrics including connectivity, flow balance, and spatial characteristics:

```{r indicators}
# Calculate comprehensive mobility indicators
indicators <- calculate_mobility_indicators(mobility_data, districts)

# Display key indicators
print("Sample of mobility indicators:")
print(head(indicators[, c("id_origin", "total_outflow", "total_inflow", 
                         "containment", "connectivity_index", "net_flow")]))

# Analyze mobility patterns
print("Mobility pattern summary:")
print(paste("Zones with net outflow (more departures):", sum(indicators$net_flow < 0, na.rm = TRUE)))
print(paste("Zones with net inflow (more arrivals):", sum(indicators$net_flow > 0, na.rm = TRUE)))
print(paste("Mean connectivity index:", round(mean(indicators$connectivity_index, na.rm = TRUE), 3)))
```

### Anomaly Detection

Detect unusual mobility patterns that deviate from normal behavior (useful for identifying special events, holidays, disruptions):

```{r anomalies}
# Detect anomalies using z-score method (accounts for weekday/weekend differences)
anomalies <- detect_mobility_anomalies(mobility_data, method = "zscore", threshold = 2, by_weekday = TRUE)

# Show detected anomalies
print("Detected mobility anomalies:")
if(any(anomalies$is_anomaly)) {
  anomaly_dates <- anomalies[anomalies$is_anomaly, ]
  print(anomaly_dates[, c("date", "total_trips", "z_score", "is_anomaly", "weekday")])
  
  # Check if anomalies correspond to holidays
  anomaly_holidays <- check_spanish_holidays(anomaly_dates$date)
  holiday_anomalies <- anomaly_holidays[anomaly_holidays$is_likely_holiday, ]
  if(nrow(holiday_anomalies) > 0) {
    print("Anomalies corresponding to holidays:")
    print(holiday_anomalies)
  }
} else {
  print("No significant anomalies detected in this period")
}

# Compare weekend vs weekday patterns
print("Weekend vs Weekday mobility patterns:")
weekend_summary <- anomalies %>%
  dplyr::group_by(is_weekend) %>%
  dplyr::summarise(
    avg_trips = mean(total_trips, na.rm = TRUE),
    anomaly_rate = mean(is_anomaly, na.rm = TRUE),
    .groups = "drop"
  )
print(weekend_summary)
```

### Distance-Decay Modeling

Model how mobility flows decrease with distance (fundamental principle in spatial interaction modeling):

```{r distance_decay}
# Calculate distance-decay relationship
# This models how trip frequency relates to geographic distance
decay_model <- calculate_distance_decay(mobility_data, districts)

# Display model results
print("Distance-decay model summary:")
print(decay_model$model_summary)
print(paste("R-squared:", round(decay_model$r_squared, 3)))
print(paste("Distance decay exponent:", round(decay_model$decay_exponent, 3)))

# Interpretation: Higher negative exponent = stronger distance effect
# Typical values range from -0.5 to -2.0 for human mobility
```

## Visualization

### Time Series Plots

Visualize mobility trends over time:

```{r daily_mobility}
daily_plot <- plot_daily_mobility(mobility_data)
print(daily_plot)
```

### Interactive Flow Maps

Create interactive maps of mobility flows:

```{r flow_maps}
# Create flow map with minimum flow threshold
flow_map <- create_flow_map(districts, mobility_data, min_flow = 500)
flow_map
```

### Choropleth Maps

Create maps showing spatial indicators:

```{r choropleth}
choropleth <- create_choropleth_map(
  districts, 
  indicators, 
  variable = "containment"
)
choropleth
```

### Heatmaps

Create flow matrix heatmaps:

```{r heatmaps}
heatmap <- plot_mobility_heatmap(mobility_data, top_n = 20)
print(heatmap)
```

### Distance-Decay Plots

Visualize distance-decay relationships:

```{r distance_plots}
decay_plot <- plot_distance_decay(decay_model)
print(decay_plot)
```

## Configuration

The package supports advanced configuration:

```{r config}
# Enable parallel processing
configure_mobspain(parallel = TRUE, n_cores = 4)

# Check configuration
mobspain_status()
```

## Conclusion

The `mobspain` package provides a complete toolkit for Spanish mobility analysis, from data retrieval to advanced analytics and visualization. 

### Key Methodological Notes

When working with MITMA data, consider:

- **Data Quality**: Mobile phone data represents a sample of the population; results should be interpreted as relative patterns rather than absolute counts
- **Temporal Patterns**: Weekdays vs. weekends show different mobility patterns; holidays and special events may create anomalies
- **Spatial Resolution**: District-level data provides fine detail but requires more computational resources; municipality or LUA levels may be more appropriate for regional studies
- **Privacy Protection**: All data is aggregated and anonymized according to Spanish data protection standards

### Further Resources

- **MITMA Methodology**: [Official methodological report](https://www.transportes.gob.es/recursos_mfom/paginabasica/recursos/a3_informe_metodologico_estudio_movilidad_mitms_v8.pdf)
- **Package Documentation**: `?mobspain` or `help(package = "mobspain")`
- **Example Scripts**: See `complete_working_example.R` and `simple_mobspain_example.R` in the package repository

For more information, see the package documentation and example scripts.
